Prospective testing for the prevalence or transience of a shock effect before it occurs

### 2021-11-18 prevalence_testing/postshocktest2021draft.tex

\begin{prop}
	Let $\bf{y}_{i,t}$, $i = 1,\ldots,n_i$, $t = 1,\ldots, T$ be a collection of independent time series. Let each time series $\bf{y}_{i,t}$ be generated by model \eqref{model-gen} with accompanying random effects structure \eqref{random-effects}, where $y_{i,t}$ is a scalar response $\x_{i,t} \in \R^p$ is a fixed vector of covariates. Suppose that there is a common shock time $p + 2 < T^* < T - (p+2)$ where $q_1 = q_2 = 1$, $\|\x_{i,t}\| = 1$, each entry of $\x_{i,t}$ is non-negative, $\mu_{\min} > 0$, and $y_{i,0} = 0$. Let $i' \geq 2$ be an index. Then, for any $\epsilon > 0$ we can choose an $\gamma > 0$ such that 
	$$
	  \left\vert\E(y_{i,t} - y_{i',t})\right\vert \leq \epsilon, \qquad |\Var(y_{1,t}) - \Var(y_{i',t})| < \varepsilon,
	$$
	provided that 
	\begin{align*}
  |\mu_\delta'\x_{1,T^*+1} - \mu_\delta'\x_{i',T^*+1}| &< \gamma, \\
%  |\mu_{\tilde\theta_1}'\x_{1,T^*+1} - \mu_{\tilde\theta_{i'}}'\x_{i',T^*+1}| &< \gamma, \\
  \left\vert\mu_{\tilde\theta_{1}}'\x_{1,t} - \mu_{\tilde\theta_{i'}}'\x_{i',t}\right\vert &< \gamma, \qquad \text{for all} \; t, \\
  |\mu_{\tilde\phi_1}^j - \mu_{\tilde\phi_{i'}}^j| &< \gamma, \qquad \text{for all} \; j, \\
  |\x_{1,T^*+1}'\Sigma_\delta\x_{1,T^*+1} - \x_{i',T^*+1}'\Sigma_\delta\x_{i',T^*+1}| &< \gamma, \\
  |\x_{1,t}'\Sigma_{\tilde\theta_{1}}\x_{1,t} - \x_{i',t}'\Sigma_{\tilde\theta_{i'}}\x_{i',t}| &< \gamma, \qquad \text{for all} \; t.
	\end{align*}
\end{prop}

@@

\noindent\textbf{This needs some work (ignore for now)} Now consider a case where there exists a $2 \leq i' < n + 1$ where $\x_{1,t} = \x_{i',t}$. Now suppose for simplicity that $T^*_{j'} = T^*_1$ and that $T_{i'} = T_1$ where $T_1 > T^*_1 + q_1 + q_2 + 2$. This assumes that we have data for the first time series beyond the shock point. We will show that when there is a covariate clone of the time series under study, then estimation in this series will approximate estimation in the time series under study provided that the underlying variability is very small relative to the signal. To see this, consider $t > T^*_1$, where 
\begin{align*}
y_{1,t} &= \eta_1 + \mu_\alpha + \delta_1'\x_{1,T^*_1+1} + \sum_{j=1}^{q_1} (\phi_{1,j} + \tilde{\phi}_{1,j})y_{1,t-j} + \sum_{j=0}^{q_2-1}(\theta_{1,j+1} + \tilde{\theta}_{1,j+1})'\x_{1,t-j} + \tilde{\varepsilon}_{1} + \varepsilon_{1,t}, \\
y_{i',t} &= \eta_{i'} + \mu_\alpha + \delta_{i'}'\x_{1,T^*_1+1} + \sum_{j=1}^{q_1} (\phi_{i',j} + \tilde{\phi}_{i',j})y_{i',t-j} + \sum_{j=0}^{q_2-1}(\theta_{i',j+1} + \tilde{\theta}_{i',j+1})'\x_{1,t-j} + \tilde{\varepsilon}_{i'} + \varepsilon_{i',t}, \\
\hat y_{1,t} &= \hat \eta_1 + \hat{\alpha}_1(\x_{1,T^*_1+1}) + \sum_{j=1}^{q_1} (\hat{\phi}_{1,j} + \hat{\tilde{\phi}}_{1,j})\hat y_{1,t-j} + \sum_{j=0}^{q_2-1}(\hat\theta_{1,j+1} + \hat{\tilde{\theta}}_{1,j+1})'\hat\x_{1,t-j}, \\
\hat y_{i',t} &= \hat\eta_{i'} + \hat{\alpha}_{i'}(\x_{1,T^*_1+1}) + \sum_{j=1}^{q_1} (\hat{\phi}_{i',j} + \hat{\tilde{\phi}}_{i',j})\hat y_{i',t-j} + \sum_{j=0}^{q_2-1}(\hat\theta_{i',j+1} + \hat{\tilde{\theta}}_{i',j+1})'\hat\x_{1,t-j}, 
\end{align*}
where $\alpha_j(\x) = \mu_\alpha + \delta_j'\x$, and we have 
\begin{align*}
  |y_{1,t} - \hat y_{1,t}| &= \Bigg\vert (\eta_1 - \hat\eta_1) + (\alpha_1 - \hat{\alpha}_1(\x_{1,T^*_1+1})) + \sum_{j=1}^{q_1}\left[(\phi_{1,j} + \tilde{\phi}_{1,j})y_{1,t-j} - (\hat\phi_{1,j} + \hat{\tilde{\phi}}_{1,j})\hat{y}_{1,t-j}\right] \\ 
  &\qquad+ \sum_{j=0}^{q_2-1}\left[(\theta_{1,j+1} + \tilde{\theta}_{1,j+1})'\x_{1,t-j} - (\hat\theta_{1,j+1} + \hat{\tilde{\theta}}_{1,j+1})'\hat\x_{1,t-j}\right] + \varepsilon_{1,t} \Bigg\vert \\
  |y_{i',t} - \hat y_{i',t}| &= \Bigg\vert (\eta_{i'} - \hat\eta_{i'}) + (\alpha_{i'} - \hat{\alpha}_{i'}(\x_{1,T^*_1+1})) + \sum_{j=1}^{q_1}\left[(\phi_{i',j} + \tilde{\phi}_{i',j})y_{i',t-j} - (\hat\phi_{i',j} + \hat{\tilde{\phi}}_{i',j})\hat{y}_{i',t-j}\right] \\ 
  &\qquad+ \sum_{j=0}^{q_2-1}\left[(\theta_{i',j+1} + \tilde{\theta}_{i',j+1})'\x_{1,t-j} - (\hat\theta_{i',j+1} + \hat{\tilde{\theta}}_{i',j+1})'\hat\x_{1,t-j}\right] + \varepsilon_{i',t} \Bigg\vert.
\end{align*}
\textbf{Needs finishing, but the argument is simple: estimation uncertainty declines as error variance decreases. And differences in the random effects decrease as variability decreases.} \vspace*{0.5cm}

%\begin{lem}[Technical Lemma 2] \label{lem1}
%	Let $a_j, b_j \in \R$, $j = 1,2$, and $C > 0$ be such that $C \leq a_j/b_j \leq 2C$, $j = 1,2$. Further suppose that 
%	$$
%	  \frac{|a_1 - a_2|}{a_1 + a_2} < \frac{\varepsilon}{2C},
%	$$
%	for some $\varepsilon > 0$. Then 
%	$$
%	  \frac{|a_1 - a_2|}{b_1 + b_2} < \varepsilon.
%	$$
%\end{lem}

%\begin{proof}
%Observe that 
%$$
%  |a_1 - a_2| < \frac{\varepsilon(a_1 + a_2)}{2C} 
%	  = \frac{\varepsilon(b_1(a_1/b_1) + b_2(a_2/b_2))}{2C} 
%	  \leq \frac{\varepsilon(2b_1C + 2b_2C)}{2C} 
%	  = \varepsilon(b_1 + b_2).
%$$
%The conclusion immediately follows from the above.
%\end{proof}

%\begin{prop} \label{prop:signaltonoisediff}
%	Suppose the conditions of Proposition~\ref{prop:signaltonoise} where $C$ is chosen so that 
%	\begin{equation} \label{signal-to-noise2}
%	  C \leq \frac{\E(y_{i,t})}{\Var(y_{i,t})} \leq 2C,
%	\end{equation}
%	for all $i$ and $1 \leq t \leq T$. Let $2 < i' < n + 1$ be the index of a time series in the donor pool for which CONDITIONS then for some $\varepsilon > 0$ we have 
%	$$
%	  \frac{\left|\E(y_{i,t} - y_{i',t})\right|}{\Var(y_{i,t} - y_{i',t})} < \varepsilon
%	$$
%\end{prop}

%\begin{proof}
%	First suppose that we have 
%	\begin{equation} \label{prop:cond}
%	  \frac{\left|\E(y_{i,t} - y_{i',t})\right|}{\E(y_{i,t} + y_{i',t})} \leq \frac{\varepsilon}{2C}.		
%	\end{equation}
%	Then Lemma~\ref{lem:absdiff} combined with the assumed conditions imply the conclusion of this proposition. We now verify that \eqref{prop:cond} holds.
%Suppose that:
%\begin{align*}
%  |\mu_\delta'\x_{1,T^*+1} - \mu_\delta'\x_{i',T^*+1}| &< \gamma \\
%  |\mu_{\tilde\theta_1}'\x_{1,T^*+1} - \mu_{\tilde\theta_{i'}}'\x_{i',T^*+1}| &< \gamma \\
%  \left\vert\mu_{\tilde\theta_{1}}'\x_{1,t} - \mu_{\tilde\theta_{i'}}'\x_{i',t}\right\vert &< \gamma, \qquad \text{for all} \; t \\
%  |\mu_{\tilde\phi_1}^j - \mu_{\tilde\phi_{i'}}^j| &< \gamma, \qquad \text{for all} \; j \\
%  \mu_\delta'\x_{1,T^*+1} + \mu_\delta'\x_{i',T^*+1} &> D
%\end{align*}
%Similarly, 
%\begin{align*}
%	&\E(y_{i,t} + y_{i',t}) = \sum_{j=0}^{t - (T^*+1)}\left[\left(\mu_{\tilde\phi_{i}}^j(\mu_\delta'\x_{i,T^*+1}) + \mu_{\tilde\phi_{i'}}^j(\mu_\delta'\x_{i',T^*+1})\right) + \mu_{\tilde\phi_{i}}^j(\mu_{\tilde\theta_{i}}'\x_{i,T^*+1}) + \mu_{\tilde\phi_{i'}}^j(\mu_{\tilde\theta_{i'}}'\x_{i',T^*+1})\right] \\
%		&\qquad + \sum_{j=0}^{T^* - 1}\left[\mu_\eta(\mu_{\tilde\phi_{i}}^{t - T^*} + \mu_{\tilde\phi_{i'}}^{t - T^*})\mu_{\phi}^j + \left(\mu_{\tilde\phi_{i}}^{t - T^*}(\mu_{\theta}'\x_{i,t-j}) + \mu_{\tilde\phi_{i'}}^{t - T^*}(\mu_{\theta}'\x_{i,t-j})\right)\mu_{\phi}^j\right].
%\end{align*}
%\end{proof}

%Let 
%$$
%  \rho_\text{min} = \min\left\{
%    \frac{\mu_\eta}{\sigma^2_\eta}, 
%    \frac{\|\mu_{\delta}\|}{\|\Sigma_\delta\|},
%    \frac{\|\mu_{\theta}\|}{\|\Sigma_\theta\|},
%    \frac{\|\mu_{\phi}\|}{\|\Sigma_\phi\|},
%    \frac{\|\mu_{\tilde\theta}\|}{\|\Sigma_{\tilde{\theta}}\|},
%    \frac{\|\mu_{\tilde\phi}\|}{\|\Sigma_{\tilde{\phi}}\|}             
%  \right\},
%$$
%and let
%$$
%  \rho_\text{max} = \max\left\{
%    \frac{\mu_\eta}{\sigma^2_\eta}, 
%    \frac{\|\mu_{\delta}\|}{\|\Sigma_\delta\|},
%    \frac{\|\mu_{\theta}\|}{\|\Sigma_\theta\|},
%    \frac{\|\mu_{\phi}\|}{\|\Sigma_\phi\|},
%    \frac{\|\mu_{\tilde\theta}\|}{\|\Sigma_{\tilde{\theta}}\|},
%    \frac{\|\mu_{\tilde\phi}\|}{\|\Sigma_{\tilde{\phi}}\|}             
%  \right\}.
%$$

>>>>

\begin{prop} \label{prop:donor}
	Let $\bf{y}_{i,t}$, $i = 1,\ldots,n_i$, $t = 1,\ldots, T$ be a collection of independent time series. Let each time series $\bf{y}_{i,t}$ be generated by model \eqref{model-gen} with accompanying random effects structure \eqref{random-effects}, where $y_{i,t}$ is a scalar response $\x_{i,t} \in \R^p$ is a fixed vector of covariates. Suppose that there is a common shock time $p + 2 < T^* < T - (p+2)$ where $q_1 = q_2 = 1$, $\|\x_{i,t}\| = 1$, each entry of $\x_{i,t}$ is non-negative, $\mu_{\min} > 0$, and $y_{i,0} = 0$. Let $i' \geq 2$ be an index. Then, for any $\epsilon > 0$ we can choose an $\gamma > 0$ such that 
	$$
	  \left\vert\E(y_{i,t} - y_{i',t})\right\vert \leq \epsilon, \qquad |\Var(y_{1,t}) - \Var(y_{i',t})| < \varepsilon,
	$$
	provided that 
	\begin{align*}
  |\mu_\delta'\x_{1,T^*+1} - \mu_\delta'\x_{i',T^*+1}| &< \gamma, \\
%  |\mu_{\tilde\theta_1}'\x_{1,T^*+1} - \mu_{\tilde\theta_{i'}}'\x_{i',T^*+1}| &< \gamma, \\
  \left\vert\mu_{\tilde\theta_{1}}'\x_{1,t} - \mu_{\tilde\theta_{i'}}'\x_{i',t}\right\vert &< \gamma, \qquad \text{for all} \; t, \\
  |\mu_{\tilde\phi_1}^j - \mu_{\tilde\phi_{i'}}^j| &< \gamma, \qquad \text{for all} \; j, \\
  |\x_{1,T^*+1}'\Sigma_\delta\x_{1,T^*+1} - \x_{i',T^*+1}'\Sigma_\delta\x_{i',T^*+1}| &< \gamma, \\
  |\x_{1,t}'\Sigma_{\tilde\theta_{1}}\x_{1,t} - \x_{i',t}'\Sigma_{\tilde\theta_{i'}}\x_{i',t}| &< \gamma, \qquad \text{for all} \; t.
	\end{align*}
\end{prop}

@@

\begin{cor} \label{cor:differences}
	Let $\bf{y}_{i,t}$, $i = 1,\ldots,n_i$, $t = 1,\ldots, T$ be a collection of independent time series. Let each time series $\bf{y}_{i,t}$ be generated by model \eqref{model-gen} with accompanying random effects structure \eqref{random-effects}, where $y_{i,t}$ is a scalar response $\x_{i,t} \in \R^p$ is a fixed vector of covariates. Suppose that there is a common shock time $p + 2 < T^* < T - (p+2)$ where $q_1 = q_2 = 1$, $\|\x_{i,t}\| = 1$, each entry of $\x_{i,t}$ is non-negative, $\mu_{\min} > 0$, and $y_{i,0} = 0$. Let $i' \geq 2$ be an index. Assume that $\Var(y_{1,t})$, $\Var(y_{i',t}) \geq b$ for all $t$.  Assume the conditions of Proposition~\ref{prop:donor}. Then we can choose an $\varepsilon > 0$ such that 
	$$
	  \left\vert \frac{\E(y_{1,t})}{\Var(y_{1,t})} - \frac{\E(y_{i',t})}{\Var(y_{i',t})} \right\vert \leq \varepsilon,
	$$
	for all $1 \leq t \leq T$.
\end{cor}

\begin{proof}
	From Proposition~\ref{prop:donor} we can pick an arbitrary number $\delta > 0$ such that $\left\vert\E(y_{i,t} - y_{i',t})\right\vert \leq \delta$ and $|\Var(y_{1,t}) - \Var(y_{i',t})| < \delta$ for all $t = 1,\ldots,T$. The conclusion follows from Lemma~\ref{lem:diffsig2noise}.
\end{proof}

This is some justification for the idea that one may be able to make inferences about the time series of interest using a time series from the donor pool.

\noindent \textbf{Doodles}:
\begin{align*}
  &\Prob\left(\left\vert |y_{1,t} - \E(y_{1,t})| - |y_{i',t} - \E(y_{i',t})| \right\vert \leq \varepsilon_1\right) 
  \geq \Prob\left(|y_{1,t} - \E(y_{1,t})| \leq \varepsilon_1/2,  |y_{i',t} - \E(y_{i',t})| \leq \varepsilon_1/2\right) \\
  &\qquad = \Prob\left(|y_{1,t} - \E(y_{1,t})| \leq \varepsilon_1/2\right) \Prob\left(|y_{i',t} - \E(y_{i',t})| \leq \varepsilon_1/2\right) \\
  &\qquad \geq \left(1 - \frac{4\Var(y_{1,t})}{\varepsilon_1^2}\right)\left(1 - \frac{4\Var(y_{i',t})}{\varepsilon_1^2}\right) \\
  &\qquad \geq \left(1 - \frac{4\E(y_{1,t})}{\varepsilon_1^2C}\right)\left(1 - \frac{4\E(y_{i',t})}{\varepsilon_1^2C}\right),
  %&\qquad = 1 - \frac{4\E(y_{1,t})}{\varepsilon^2C} - \frac{4\E(y_{i',t})}{\varepsilon^2C} + \frac{16\E(y_{1,t})\E(y_{i',t})}{\varepsilon^4C^2},
\end{align*}
for all $t = 1,\ldots,T$ where the second inequality follows from Chebyshev's inequality and the last inequality follows from Proposition~\ref{prop:signaltonoise}.

\vspace*{0.5cm}
The above argument implies that a $C$ can be chosen so that 
\begin{align*}
  &|y_{1,t} - y_{i',t}| = |y_{1,t} - y_{i',t} \pm \E(y_{1,t}) \pm \E(y_{i',t})| \\
  &\qquad = |(y_{1,t} - \E(y_{1,t})) - (y_{i',t} - \E(y_{i',t})) + (\E(y_{1,t}) - \E(y_{i',t}))| \\
  &\qquad \leq |y_{1,t} - \E(y_{1,t})| + |y_{i',t} - \E(y_{i',t})| + |\E(y_{1,t}) - \E(y_{i',t})| \\
  &\qquad \leq \varepsilon_2,
\end{align*}
holds with high probability for all $t = 1,\ldots,T$ by Proposition~\ref{prop:donor}.

\vspace*{0.5cm}
Add condition that $\text{argmax}_t\|\x_{1,t} - \x_{i',t}\| < \delta$. Then the OLS estimates satisfy 
$$
  \|(\hat\eta_1,\hat\phi_1,\hat\theta_1',\hat\alpha_1)' - 
    (\hat\eta_{i'},\hat\phi_{i'},\hat\theta_{i'}',\hat\alpha_{i'})'\| < \varepsilon_3
$$
with high probability, and therefore $\left\vert|y_{1,t} - \hat y_{i,t}| - |y_{i',t} - \hat y_{i',t}|\right\vert < \varepsilon_4$ with high probability for all $t = 1,\ldots,T$.
